---
title: "Mini Data Analysis Milestone 2"
output: html_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
suppressMessages(library(datateachr)) # <- might contain the data you picked!
suppressMessages(library(tidyverse))
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Which types of trees have been planted the most in the past five years in each neighborhood? In other words, is there a trend in cultivated tree species in each area?
2. Is there any kind of relation between the height of trees (with the same type) and the neighborhood they have been planted in?
3. Within each neighborhood, do trees with a curb around them tend to have smaller diameters (in terms of growth)?
4. Could we validate within the dataset that the age of a tree directly relates to its diameter? Is it the same rate among all tree species, or it differs from one to another?
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
#### Question 1: any trend in cultivated tree species in each area in past 5 years?

For this question, I choose **computing number of observations (option 2)** for the summarizing task. If we are going to compare between each area that which tree species is most popular, it would be suitable to know exactly how many trees were planted that year in neighborhoods. For example, if we're saying 80% percent of planted trees were from ABIES in Kits and 70% in Downtown, the number of total planted trees would be of interest to be more accurate. Also, it is useful in comparison between tree species in different years.

```{r}
num_tr_nghbr <- vancouver_trees %>%
  # creating a variable year when trees are planted
  mutate(year_planted = as.character(substring(date_planted, 1, 4))) %>%
  # grouping trees by their neighborhood & year
  group_by(neighbourhood_name, year_planted) %>%
  # trees has planted in the pas 10 years
  filter(as.numeric(substring(Sys.Date(), 1, 4)) - 
          as.numeric(year_planted) <= 5) %>%
  # counting number of trees per year in each neighborhood
  tally()

num_tr_nghbr
```

I choose **option 6 (logarithmic scale and formatting the axes labels)** for this question. The plot is trying to convey the summarized variable in a way that is easy to compare the number of planted trees across years and neighborhoods.

```{r}
hist_tr_nghbr <- num_tr_nghbr %>%
  ggplot(aes(neighbourhood_name, n)) +
  geom_col(aes(fill=year_planted), position = "dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_log10() +
  xlab("Neighbourhood") +
  ylab("Number of trees (log)")

print(hist_tr_nghbr)
```

#### Question 2: Any relation between the height of trees (with the same type) and the neighborhood?

I choose **option 1: summary statistics** for this question. I calculated mean, min, max, median, and standard deviation of height across neighborhoods and tree species. It would be useful to know the mean and other statistics on the height when we're comparing it between two neighborhoods.

```{r}
height_nghbr <- vancouver_trees %>%
  group_by(neighbourhood_name, genus_name) %>%
  summarise(mean_height= mean(height_range_id),
            min_height = min(height_range_id),
            max_height =  max(height_range_id),
            median_height = median(height_range_id),
            std_height = sd(height_range_id), .groups = 'drop')  #%>%

  #filter(genus_name == "PRUNUS")

height_nghbr
```
Since we have many types of tree species in the data, I focused only on the family of trees instead of species and considered the top 3 that are most populated in all neighborhoods to investigate the question.  I plotted the density distribution of height of the tree families across neighborhoods (Implementing **options 7**).

```{r}
top_genus <- vancouver_trees %>%
  group_by(genus_name) %>%
  tally() %>%
  arrange(desc(n)) %>%
  top_n(3)

height_dist_nghbr <- vancouver_trees %>%
  filter(genus_name %in% top_genus$genus_name) %>%
  group_by(genus_name) %>%
  ggplot(aes(height_range_id)) +
  geom_density(aes(fill=genus_name), alpha = 0.5) +
  facet_wrap(~ neighbourhood_name)
  
height_dist_nghbr
```


It seems the high distributions vary for each tree family across neighborhoods! As shown, the trees in the ACER family are higher in some neighborhoods than others. But, the distribution of PRUNUS does not vary much within neighborhoods.

#### Question 3: curb affects diameters?

For this research question, It would be good to know how many trees have curbs around them and how many do not. Because it would not be a fair comparison if we have a lot of observations from one population and few samples from another one. It turns out to be a case in our question! we do have fewer samples for ``curb=NO`` against ``curb=YES`` in each neighborhood. 

```{r}
count_curbed <- vancouver_trees %>%
  group_by(neighbourhood_name, curb) %>%
  tally()

count_curbed
```

I plotted the distribution of trees diameter with and without curbs in each neighborhood. so that it can help us identify does having a curb affects the tree's diameter. As shown in the plot, there is no significant difference between the distributions and their means  in most neighborhoods.

```{r}
means <- vancouver_trees %>%
  filter(0 < diameter & diameter < 50) %>%
  group_by(neighbourhood_name, curb) %>%
  summarise(mean_dmtr = mean(diameter), .groups = 'drop')

dist_dmtr_nghbr <- vancouver_trees %>%
  filter(0 < diameter & diameter < 50) %>%
  group_by(neighbourhood_name) %>%
  ggplot(aes(diameter)) + 
  geom_density(aes(fill=curb), alpha=0.5) +
  geom_vline(data=means, aes(xintercept=means$mean_dmtr, color=curb), 
             linetype="dashed") +
  facet_wrap(~ neighbourhood_name)

dist_dmtr_nghbr
```

#### Question 4: Age Vs diameter across tree species?

I created a categorical variable, named age_cat, based on the age of trees. This new variable has five categories: ``Youth, Mature, Middle Age, Senior, Ancient``. The reason for creating this was to first, have better visualization of age against diameter as scatter plot for this question was not that informative, and second, investigate how diameter is distributed in each period of ages.

```{r}
van_tree_catAge <- vancouver_trees %>%
  mutate(age = as.numeric(substring(Sys.Date(), 1, 4))-
                 as.numeric(substring(date_planted, 1, 4))) %>%
  mutate(age_cat = cut(age, 
                breaks = c(0, 10, 20, 30, 40, Inf), 
                labels = c("Youth","Mature","Middle Age","Senior","Ancient"))) %>%
  select(tree_id, age, age_cat, everything())

head(van_tree_catAge)
```

For the top 3 tree species that we have the most data on, I tried to explore how statistics of diameters change among the period of ages. I plotted a box plot of diameters for each age group and compared them. As you can see, for each species, the median, upper and lower quantile grows as we go forward in age periods, meaning that there is a positive correlation between diameter and age of trees!

```{r}
age_diamtr_rel <- van_tree_catAge %>%
  filter(genus_name %in% top_genus$genus_name) %>%
  filter(0 < diameter & diameter < 100) %>%
  filter(!is.na(diameter) & !is.na(age)) %>%
  ggplot(aes(age_cat,  diameter)) +
  geom_boxplot(aes(fill=age_cat)) +
  theme(axis.text.x = element_text(color = "black", size=11, angle=30, vjust=.8, hjust=0.8)) +
  scale_y_log10() +
  facet_wrap(~ genus_name)

age_diamtr_rel
```



<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
I think for question 1, so far I get the background knowledge for it and I need to involve tree species information to be closer to the answer. I think adding one layer of tree species grouping in each year to the provided plot would reveal a lot of information about the question, but it needs additional work in terms of plotting to better convey the results.

With the operations we did (statistics of height and distributions), we can say there are not any specific differences for some types of trees like the PRUNUS among neighborhoods (3 modalities). But, I'd like to explore in more depth species like ACER. Another thing to consider is the amount of data that we have for each species in each neighborhood. For example,  if only we have a handful number of trees in a neighborhood, it would be statistically significant we draw a conclusion based on those numbers of trees. And,  that is why I tried to narrow my focus on tree species that we have more data.

For question 3, we have an unbalanced number of observations for each class of curb, and it is problematic to draw a conclusion based on. There are some subtle differences in distributions of diameter for each class of curb in each neighborhood but that is not that high. Also, I think we should consider the age of trees in diameter and only compare trees with the same age in further analysis. Because older trees tend to have bigger diameters and this fact makes our analysis biased.

Based on the provided box plot, we're confident to say that there is a positive correlation between the age of trees and their diameters for the explored tree species. But what I am curious about is to fit a linear model on age versus diameter scatter-plot and see how coefficients differ from one to another and which one's diameter grows more when it gets older.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

Let's start with looking at the data again  and  see if it is satisfied with our criteria of a tidy data.

```{r}
glimpse(vancouver_trees)
```

As we can see, each row is an observation unit of trees, and each variable is a record from them. As far as my research questions were, we didn't need any re-formatting (wide to long and vice versa) to make our analysis easier and more efficient since we could put the names of our variables to the axis of plots. Another criteria that the data set meets is that each cell represents a value. So, our data is indeed tidy.

<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

First, start with untidy-ing data. For this purpose, I combine ``diameter`` and ``height_range_id`` variables together as ``dimension_name`` and their values to ``dimension_value`` to make a long version of data. As shown below, now the dimension_name variable is no longer a record of our data and it doesn't represent a value, which makes it untidy! With the below dataset, answering questions 3 and 4 would be hard.

```{r}
untidey_data <- vancouver_trees %>%
  pivot_longer(
    cols = c(diameter, height_range_id),
    names_to = "dimension_name",
    values_to = "dimension_value"
  ) %>%
  select(tree_id, dimension_name, dimension_value, everything())

head(untidey_data)
```

Now let's try to make the data back to tidy again!

```{r}
tidy_data <- untidey_data %>%
  pivot_wider(
    id_cols = c(-dimension_name, -dimension_value),
    names_from = dimension_name,
    values_from = dimension_value) %>%
  select(tree_id, diameter, height_range_id, everything())
head(tidy_data)
```

Our data is tidy again!

<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->

Based on what I have done so far in research questions, I would like to continue my work on question 1 (trends in cultivated tree species in each neighborhood past five years) and question 4 (Age verus diameter across tree species). I think these questions are more likely to yield correct and interesting answers. Unbalanced classes of data in the two other questions are barriers to producing concrete results. Also, we somewhat reached the answers we wanted based on the summarized statistics and plots, and we could see that is not going to yield interesting insights. Besides, I am excited to fit a linear model to explore the relation between diameter and age for each species in question 4. For question 1, I am curious about how to plot the information in a way to see trends in each neighborhood.

<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 