---
title: "Mini Data Analysis Milestone 2"
output: html_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Which types of trees have been planted the most in the past ten years in each neighborhood? In other words, is there a trend in cultivated tree species in each area?
2. Is there any kind of relation between the height of trees (with the same type) and the neighborhood they have been planted in?
3. Within each neighborhood, do trees with a curb around them tend to have smaller diameters (in terms of growth)?
4. Could we validate within the dataset that the age of a tree directly relates to its diameter? Is it the same rate among all tree species, or it differs from one to another?
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
#### Question1: any trend in cultivated tree species in each area in past 10 years?

```{r}
num_tr_nghbr <- vancouver_trees %>%
  # creating a variable year when trees are planted
  mutate(year_planted = as.character(substring(date_planted, 1, 4))) %>%
  # grouping trees by their neighborhood & year
  group_by(neighbourhood_name, year_planted) %>%
  # trees has planted in the pas 10 years
  filter(as.numeric(substring(Sys.Date(), 1, 4)) - 
          as.numeric(year_planted) <= 10) %>%
  # counting number of trees per year in each neighborhood
  tally()

num_tr_nghbr
```

```{r}
hist_tr_nghbr <- num_tr_nghbr %>%
  ggplot(aes(neighbourhood_name, n)) +
  geom_col(aes(fill=year_planted)) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_brewer(palette = "Paired") +
  xlab("Neighbourhood") +
  ylab("Number of trees (log)")

print(hist_tr_nghbr)
```

#### Question2: Any relation between the height of trees (with the same type) and the neighborhood?

```{r}
height_nghbr <- vancouver_trees %>%
  group_by(neighbourhood_name, genus_name) %>%
  summarise(mean_height= mean(height_range_id),
            min_height = min(height_range_id),
            max_height =  max(height_range_id),
            median_height = median(height_range_id),
            std_height = sd(height_range_id), .groups = 'drop')

height_nghbr
```


```{r}
top_genus <- vancouver_trees %>%
  group_by(genus_name) %>%
  tally() %>%
  arrange(desc(n)) %>%
  top_n(3)

height_dist_nghbr <- vancouver_trees %>%
  filter(genus_name %in% top_genus$genus_name) %>%
  group_by(genus_name) %>%
  ggplot(aes(height_range_id)) +
  geom_density(aes(fill=genus_name), alpha = 0.5) +
  facet_wrap(~ neighbourhood_name)
  
height_dist_nghbr
```

```{r}
count_curbed <- vancouver_trees %>%
  group_by(neighbourhood_name, curb) %>%
  tally()

count_curbed
```

```{r}
means <- vancouver_trees %>%
  filter(diameter < 50) %>%
  group_by(neighbourhood_name, curb) %>%
  summarise(mean_dmtr = mean(diameter), .groups = 'drop')

dist_dmtr_nghbr <- vancouver_trees %>%
  filter(diameter < 50) %>%
  group_by(neighbourhood_name) %>%
  ggplot(aes(diameter)) + 
  geom_density(aes(fill=curb), alpha=0.6) +
  geom_vline(data=means, aes(xintercept=means$mean_dmtr, color=curb), 
             linetype="dashed") +
  facet_wrap(~ neighbourhood_name)

dist_dmtr_nghbr
```

```{r}
van_tree_catAge <- vancouver_trees %>%
  mutate(age = as.numeric(substring(Sys.Date(), 1, 4))-
                 as.numeric(substring(date_planted, 1, 4))) %>%
  mutate(age_cat = cut(age, 
                breaks = c(0, 10, 20, 30, 40, Inf), 
                labels = c("Youth","Mature","Middle Age","Senior","Ancient"))) %>%
  select(tree_id, age, age_cat, everything())

head(van_tree_catAge)
```

```{r}
age_diamtr_rel <- van_tree_catAge %>%
  filter(genus_name %in% top_genus$genus_name) %>%
  filter(0 < diameter & diameter < 100) %>%
  filter(!is.na(diameter) & !is.na(age)) %>%
  ggplot(aes(age_cat,  diameter)) +
  geom_boxplot(aes(fill=age_cat)) +
  theme(axis.text.x = element_text(color = "black", size=11, angle=30, vjust=.8, hjust=0.8)) +
  scale_y_log10() +
  facet_wrap(~ genus_name)

age_diamtr_rel
```



<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->


<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 